---
title: "Clusters and Mapping"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clusters and Mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tarr,  quietly = TRUE)
library(sf,    quietly = TRUE)
```



## 4. GIS & clustering
For spatial analysis, several base layers and functions are provided for mapping and clustering detection. The R package
"sf" is used in most spatial transformations and analysis.  Geocoded data can be mapped both statically and interactively.

The named list variable `geo_sets` stores paths to often used base layers. For example `geo_sets$isd` returns the path to
a layer of independent school districts. The next table shows the name and the layer available, all of which are polygons
stored as "sf" objects and have a coordinate reference system set `epsg:4326`

|Name             |Layer                                       |
|:--------------- |:-------------------------------------------|
|ZCTA             |Tarrant Zip Code Tabulation Areas           |
|tarrant_border   |Tarrant County boundaries as a polygon      |
|tarrant_tracts   |Census tracts in Tarrant County             |
|tarrant_blocks   |Census blocks in Tarrant County             |
|tarrant_cities   |Boundaries of partially or wholly contained cities|
|surrounding_borders|Borders of counties surrounding Tarrant   |
|isd              |Independent School Districts                |

Each of the layers can be loaded using `load_tarrant_spatial()` or `readRDS()`.  The load_tarrant_spatial(.name, .crs)
just takes one of the names in the table above and by default returns the layer by default with a Coordinate Reference
System (CRS) of '4326'. You can also use the base R function readRDS() and give it the path to the layer using the
geo_sets list varibale.  Here is how each method can be used to load a layer.

```{r reading_layers}
# load and add city boundaries using readRDS and geo_sets
other_counties <- readRDS(file = geo_sets$surrounding_borders)
plot(x = st_geometry(other_counties), border = "grey60")

# load the Tarrant County border then add it to the map
tarr_border <- load_tarrant_spatial("tarrant_border")
plot(x = st_geometry(tarr_border), border = "red", add = TRUE )

rm(cities)

```

Density based algorithms are provided for idnetifying possible clusters.  

**Key functions**: `scan_cluster()` (DBSCAN/OPTICS/HDBSCAN), `autoplot.clustered()`, `plot.clustered()`, `scan_object()`, `get_unit()`, `load_tarrant_spatial()`, `map_theme`.

```r
# Optional basemap layers

spat <- load_tarrant_spatial()

# Ensure geometry 
# cases_sf <- sf::st_as_sf(nbs, coords = c(lon,lat), crs = 4326)

# Cluster (example: HDBSCAN)
cl <- scan_cluster(
  x = cases_sf,
  method = hdbscan,
  minPts = 5
)

# Visualize (ggplot)
autoplot(cl) + map_theme

# Or with tmap
plot(cl)

# Inspect clustering internals
scan_object(cl)
get_unit(cl)
```

**Tips**
- Prefer a projected CRS for distance-based clustering (choose one appropriate for your county/state).
- HDBSCAN handles variable density better; DBSCAN is fast for well-separated clusters.

---

