---
title: "Reading and Combining EpiTrax and NBS Data"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Reading and Combining EpiTrax and NBS Data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
library(tarr, quietly = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
```

## Overview

The `tarr` package provides separate functions for reading EpiTrax and
NBS exports because the two systems use different field names, report
formats, and value conventions. The general process is:

1. Export the appropriate CSV report from EpiTrax or NBS.
2. Import and process each report with its system-specific function.
3. Convert the processed EpiTrax data to the NBS-compatible schema.
4. Combine records and remove duplicates using the intended workflow.

The first part of this vignette explains what each public function does.
The second part shows how the functions fit together in routine
workflows.

### Protected health information

> **Warning:** EpiTrax and NBS exports may contain protected health
> information (PHI). Use real exports only within the approved secure
> county environment. Do not copy, upload, email, or process identifiable
> case data on unapproved devices or services. Outside the secure county
> environment, use only synthetic or properly de-identified data.

## Functions

| Function | Primary purpose |
|:--|:--|
| `epitrax_import()` | Read and process an EpiTrax line-list CSV |
| `nbs_import()` | Read an NBS CSV and pass it through `nbs_process()` |
| `nbs_process()` | Standardize an NBS report already loaded in R |
| `epitrax_2_nbs_format()` | Convert processed EpiTrax records to the NBS-compatible schema |
| `retrieve_base_data()` | Read the maintained NBS base Parquet file |
| `baseData()` | Combine historical NBS base data with current EpiTrax records |
| `baseNBSUpdate()` | Replace a finalized date range in the maintained NBS base |
| `construct_nbs_base()` | Rebuild the NBS base from the standard folder of NBS CSV exports |

### `epitrax_import()`

`epitrax_import()` reads and processes an EpiTrax line-list CSV. Specifically the month report export from EPiTrax.  Supply the filename with `.file`, or omit it to select the file interactively.

```{r epitrax-import}
epitrax_file <- file.path(
  paths$data,
  "EpiTrax_year_to_date.csv"
)

epitrax <- epitrax_import(
  .file = epitrax_file,
  .include.all = FALSE,
  .use_cache = FALSE
)
```

The function:

- reads all source columns as character data and cleans their names;
- verifies that the file resembles an EpiTrax line-list report;
- returns confirmed and probable cases by default;
- converts date, MMWR week, ZIP code, latitude, and longitude fields;
- adds `event_month`, `onset_mmwr_week`, and `age_yrs`; and
- checks for missing county values as a result of address at diagnosis not selected.

Set `.include.all = TRUE` to include confirmed, probable, suspect,
lost-to-follow-up, and not-a-case records. This option does not mean
that every possible status value is returned.

`epitrax_import()` does not remove duplicate records. De-duplication is
performed later when EpiTrax records are combined with the NBS base.

To inspect records with a missing county from lack of address at diagnosis, use:

```{r epitrax-missing-county}
missing_county <- epitrax_import(
  .file = epitrax_file,
  .use_cache = FALSE,
  .return_missing_county = TRUE
)
```

### `nbs_import()`

`nbs_import()` reads an NBS CSV report and passes the resulting data frame to `nbs_process()`. `nbs_import()`  function targets the NBS report of `All cases with jurisdiction security` and all columns included in the report.  


```{r nbs-import}
nbs_file <- file.path(
  paths$communicable,
  "NBS",
  "Jan_2025_to_Dec_2025.csv"
)

nbs <- nbs_import(
  .file = nbs_file,
  .include.all = FALSE,
  .use_cache = TRUE
)
```

By default, `nbs_import()` returns confirmed and probable cases. It also
uses a Parquet cache beside the source CSV. If a matching Parquet file
exists and `.use_cache = TRUE` which os the default, the cached data are returned. Set
`.use_cache = FALSE` after replacing or correcting the CSV so that the
source report is processed again.

### `nbs_process()`

`nbs_process()` processes an NBS report that has already been read into
R. Most users should call `nbs_import()`, which handles both reading and
processing. Call `nbs_process()` directly when the report is already a
data frame or when developing and validating an import workflow.

```{r nbs-process}
nbs_raw <- utils::read.csv(
  nbs_file,
  header = TRUE,
  check.names = TRUE,
  stringsAsFactors = FALSE
)

nbs <- nbs_process(
  .df = nbs_raw,
  .include.all = FALSE
)
```

The function:

- recognizes supported NBS report layouts;
- combines paired NBS reports when the selected format requires a
  second export;
- retains Tarrant County records with valid person and case-status data;
- returns confirmed and probable cases by default;
- standardizes dates, numeric fields, ZIP codes, city names, sex, race,
  ethnicity, and case-status values;
- converts recognized yes/no/unknown fields to logical vectors;
- calculates age, event month, and MMWR dates; and
- selects fields used in the common NBS/EpiTrax schema.

The `.report_type` argument bypasses automatic report classification and
is intended primarily for package development and controlled imports.

### `epitrax_2_nbs_format()`

`epitrax_2_nbs_format()` translates processed EpiTrax records to the
NBS-compatible field names and data types. Run it after
`epitrax_import()` and before combining EpiTrax and NBS records.

```{r epitrax-to-nbs}
epitrax_nbs <- epitrax_2_nbs_format(
  et = epitrax,
  nbs = nbs
)
```

When `nbs` is supplied, it acts as the schema template. The returned
EpiTrax data contain the relevant NBS columns with compatible types.
The function also maps EpiTrax condition names to their NBS equivalents
and converts recognized yes/no/unknown fields to logical vectors.

This function does not remove duplicates.

### `retrieve_base_data()`

`retrieve_base_data()` reads the processed NBS base Parquet file.  That parquet file contains 
all the probale and confirmed case records for Tarrant County since 
January 1, 2005 to the last update.  The file path to the base NBS data by default is 
`r file.path(paths$communicable, "NBS", "nbs_base.parquet")`

```{r retrieve-base}
nbs_base <- retrieve_base_data(nbs_base_file)

```

Use this function when you need only the historical NBS base without
adding current EpiTrax records.

### `baseData()`

`baseData()` is the main production function for returning historical
NBS data together with current EpiTrax records.

```{r base-data}
all_cases <- baseData(
  .base = nbs_base_file,
  .chooseNew = TRUE
)
```

With `.chooseNew = TRUE`, the function:

1. reads the NBS base Parquet file;
2. prompts for the current EpiTrax CSV;
3. processes the EpiTrax report;
4. converts EpiTrax fields to the NBS-compatible schema;
5. keeps columns shared by both sources;
6. labels each record as `"Epitrax"` or `"NBS"`; and
7. removes duplicate records.

EpiTrax records are placed before NBS records during the final
combination. When the same person, birth time, condition, and event date
occur in both sources, the EpiTrax record is retained. Use
`.chooseNew = FALSE` to return only the existing NBS base.

### Base maintenance functions

`baseNBSUpdate()` replaces an event-date range in the NBS base with a
newly exported and processed NBS report. It backs up the existing base
before writing the replacement. The selected CSV filename must follow
the expected pattern, such as `Jan_2025_to_Dec_2025.csv`.

```{r base-update}
updated_base <- baseNBSUpdate(.base = nbs_base_file)
```

`construct_nbs_base()` rebuilds `nbs_base.parquet` from the set of NBS
CSV exports in the standard NBS folder under `paths$communicable`. It is
intended for recovery, correction, or a deliberate full rebuild rather
than routine report analysis.

```{r construct-base}
construct_nbs_base()
```

Both functions overwrite the base after creating a backup or obtaining
interactive confirmation. Confirm the selected files, date ranges, and
backup location before proceeding.

## Workflows

### Read and inspect one EpiTrax export

Use this workflow when the analysis needs EpiTrax fields in their native
cleaned form and does not need to combine them with NBS data.

```{r workflow-epitrax}
epitrax <- epitrax_import(
  .file = epitrax_file,
  .include.all = FALSE,
  .use_cache = FALSE
)

nrow(epitrax)
names(epitrax)
table(epitrax$patient_lhd_case_status, useNA = "ifany")
summary(epitrax$patient_event_onset_date)
```

Review unexpected status values, missing counties, missing event dates,
and invalid coordinates before using the records in reports.

### Read and inspect one NBS export

Use this workflow for an NBS-only report or to validate an NBS export
before updating the base.

```{r workflow-nbs}
nbs <- nbs_import(
  .file = nbs_file,
  .include.all = FALSE,
  .use_cache = FALSE
)

nrow(nbs)
names(nbs)
table(nbs$Case.Status, useNA = "ifany")
range(nbs$Event.Date, na.rm = TRUE)
```

Setting `.use_cache = FALSE` is useful during validation because it
forces the selected CSV to be read and processed.

### Read and combine explicit EpiTrax and NBS files

This workflow is useful when you want complete control over both input
filenames rather than using the interactive prompt in `baseData()`.

```{r workflow-combine-explicit}
# Read and process both reports.
epitrax <- epitrax_import(
  .file = epitrax_file,
  .include.all = FALSE,
  .use_cache = FALSE
)

nbs <- nbs_import(
  .file = nbs_file,
  .include.all = FALSE,
  .use_cache = FALSE
)

# Convert EpiTrax to the NBS schema.
epitrax_nbs <- epitrax_2_nbs_format(
  et = epitrax,
  nbs = nbs
)

# Keep the same columns and identify the source system.
common_fields <- intersect(names(nbs), names(epitrax_nbs))

epitrax_nbs <- epitrax_nbs[, common_fields, drop = FALSE]
nbs <- nbs[, common_fields, drop = FALSE]

combined <- dplyr::bind_rows(
  EpiTrax = epitrax_nbs,
  NBS = nbs,
  .id = "dataset"
)

# Apply the same cross-system duplicate key used by baseData().
combined <- dplyr::distinct(
  combined,
  Person.Name,
  Birth.Time,
  Condition,
  Event.Date,
  .keep_all = TRUE
)
```

Because EpiTrax is bound first, this example retains the EpiTrax record
when the duplicate key matches in both sources. Review the duplicate key
before changing its fields or source precedence.

### Routine combined-data workflow

For routine production work, use the maintained NBS base for historical
records and select the current year-to-date EpiTrax export when prompted:

```{r workflow-routine}
nbs_base_file <- file.path(
  paths$communicable,
  "NBS",
  "nbs_base.parquet"
)

cases <- baseData(
  .base = nbs_base_file,
  .chooseNew = TRUE
)

table(cases$dataset, useNA = "ifany")
range(cases$Event.Date, na.rm = TRUE)
```

This workflow follows the package's intended division of data:

- the NBS base supplies finalized historical records; and
- the selected EpiTrax report supplies current records.

After NBS data for a period are finalized, use `baseNBSUpdate()` to
replace that period in the maintained base. Use `construct_nbs_base()`
only when the base must be rebuilt from all available NBS exports.

## Validation checklist

Before using the combined data in analysis or reporting, verify:

- the source filenames and report templates are correct;
- the expected case statuses were included;
- event dates cover the intended reporting period;
- missing diagnosis counties have been reviewed;
- condition names were translated as expected;
- the number of records removed as duplicates is plausible;
- `dataset` counts are plausible for the selected period; and
- the NBS base backup exists before any update or rebuild.

The import functions standardize known fields, but they cannot determine
whether the selected export used the correct report parameters or date
range. That validation remains part of the analyst's workflow.
