---
title: "Maps and Clusters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Maps and Clusters}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  fig.width = 8,
  fig.height = 6,
  out.width = "100%"
)
```

## Overview

This vignette introduces a practical mapping and clustering workflow
using `tarr`, `sf`, and the density-based clustering methods available
through `scan_cluster()`.

The examples use the `synthetic_outbreak` data frame included with the
`tarr` package. It contains simulated cases of two fake conditions in
Tarrant County. The same workflow can be adapted to a real line list
that has one row per case and columns for longitude and latitude.

By the end of this vignette, you should be able to:

-   load common Tarrant County spatial layers,
-   convert a line list into an `sf` POINT object,
-   plot case points on simple static maps,
-   use package-provided static base map tiles,
-   understand why coordinate reference systems matter,
-   choose a projected coordinate reference system for clustering,
-   run `scan_cluster()`, and
-   compare DBSCAN, HDBSCAN, and OPTICS results.

This vignette assumes that you are fairly new to GIS work in R. The
examples are intentionally verbose so that you can see each step.

## Packages used

```{r packages, message = FALSE, warning = FALSE}
library(tarr)
library(maptiles)
library(sf)
library(dplyr)
library(ggplot2)
library(units)
library(dbscan)
```

The most important package for GIS work is `sf`. An `sf` object is a
data frame with a special geometry column. The regular columns contain
attributes such as condition, sex, onset date, or case ID. The geometry
column contains the spatial information, such as a point location or
polygon boundary.

## Spatial function categories

The spatial tools in `tarr` support the steps epidemiologists commonly
take when preparing, analyzing, and presenting geocoded disease data.
The table groups the user-facing functions by purpose.

| Category | Functions and objects | Use in disease-data work |
|:-----------------------|:-----------------------|:-----------------------|
| Address preparation | `correct_city()`, `makeAddress()` | Standardize city names and assemble address fields before geocoding |
| Reference layers | `geo_sets`, `load_tarrant_spatial()` | Locate and load county boundaries, ZIP Code Tabulation Areas, census geography, cities, and school districts as `sf` layers |
| Point-cluster detection | `scan_cluster()` | Identify spatial concentrations of disease-case points with DBSCAN, HDBSCAN, or OPTICS |
| Cluster inspection | `scan_object()`, `scan_unit()`, `print()` | Inspect the underlying clustering result, distance unit, cluster count, and noise count |
| Cluster geometry | `get_cluster_polys()` | Create interpretive polygons around groups of clustered points |
| Mapping and presentation | `plot()`, `autoplot()`, `map_theme` | Produce interactive, base, or `ggplot2` maps of cases and detected clusters |

`scanCluster()` and `mapCluster()` support older workflows. New analysis
should use `scan_cluster()` and the `plot()` or `autoplot()` methods for
the returned `clustered` object.

Several additional functions in the source files support these public
interfaces. For example, the `clustered` class constructor, validator,
subsetting methods, CRS transformation helper, polygon helper, and
`ggmap`-to-raster converter preserve object structure or prepare data
for plotting. They are implementation details rather than the normal
entry points for an epidemiologist's analysis.

## Tarrant County spatial layers

The `tarr` package includes several commonly used Tarrant County spatial
layers. The named list `geo_sets` stores the file paths for these
layers. The helper function `load_tarrant_spatial()` loads one of these
layers as an `sf` object.

Common layers include:

| Name | Description | Geometry |
|:-----------------------|:-----------------------|:-----------------------|
| `ZCTA` | ZIP Code Tabulation Areas | Polygon |
| `tarrant_border` | Tarrant County boundary | Polygon |
| `tarrant_tracts` | Census tracts in Tarrant County | Polygon |
| `tarrant_blocks` | Census blocks in Tarrant County | Polygon |
| `tarrant_cities` | Boundaries of cities that are partially or wholly in Tarrant County | Polygon |
| `surrounding_borders` | Borders of counties surrounding Tarrant County | Polygon |
| `isd` | Independent school districts | Polygon |

You can see the available layer names with:

```{r geo-set-names}
names(geo_sets)
```

The usual way to load one layer is:

```{r load-county}
county <- load_tarrant_spatial("tarrant_border")

county
```

The result is an `sf` object. You can check the geometry type and
coordinate reference system:

```{r inspect-county}
sf::st_geometry_type(county)
sf::st_crs(county)
```

The default coordinate reference system for these layers is EPSG:4326,
which stores coordinates as longitude and latitude. You can also request
a different coordinate reference system when loading a layer. This is
useful when you want a layer to match another spatial object.

```{r load-county-web}
county_3857 <- load_tarrant_spatial("tarrant_border", .crs = 3857)

sf::st_crs(county_3857)
```

## A first static map

Before mapping cases, it helps to map the county boundary and nearby
counties.

```{r simple-boundary-map}
surrounding <- load_tarrant_spatial("surrounding_borders")

plot(st_geometry(surrounding), border = "grey60")
plot(st_geometry(county), border = "red3", lwd = 2, add = TRUE)
```

This uses base R plotting. It is quick and useful for checking that a
spatial layer loaded correctly. The same map can be made with `ggplot2`:

```{r ggplot-boundary-map}
ggplot() +
  geom_sf(data = surrounding, fill = NA, color = "grey70") +
  geom_sf(data = county, fill = NA, color = "red3", linewidth = 0.8) +
  labs(title = "Tarrant County and surrounding counties") +
  map_theme
```

The `map_theme` object is a ggplot2 theme provided by `tarr`. It removes
axis labels and grid lines that are usually not helpful in mapping local
areas and sets font characterists. .

## Making a POINT sf object from a line list

A line list is usually stored as a data frame with one row per case and
attributes for each case in columns (e.g., age, address, etc.). If the
line list has coordinates like longitude/latitude columns, it can be
converted into an `sf` POINT object.

The package `synthetic_outbreak` data frame has columns named `lon` and
`lat`.

```{r look-at-linelist}
synthetic_outbreak |>
  select(condition, onset, lon, lat) |>
  head()
```

For this vignette, we will focus on the fake condition `fooflu`.

```{r filter-fooflu}
foo_cases <- synthetic_outbreak |>
  filter(condition == "fooflu")

nrow(foo_cases)
```

Now convert the ordinary data frame foo_cases into an `sf` object that
will be called `foo`.

```{r line-list-to-sf}
foo <- foo_cases |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

foo
```

The `coords` argument identifies the coordinate columns. Longitude is
the x-coordinate and latitude is the y-coordinate, so the order is
`c("lon", "lat")`.

The `crs = 4326` argument tells `sf` that the source coordinates are
longitude and latitude in EPSG:4326. Assigning the correct source CRS is
essential. If the CRS is missing or incorrectly assigned, R cannot
reliably transform the points, measure distances, or align them with
reference layers.

You can confirm that `foo` is an `sf` POINT object and inspect its CRS.
The printed Well-Known Text is lengthy, but important fields include the
angular unit (`degree`), area of use, datum, and EPSG identifier.

```{r inspect-foo}
sf::st_geometry_type(foo)
sf::st_crs(foo)
```

## Plotting case points

A quick base R plot is often the easiest first check. Note the
st_geomtery() function. It pulls the geometry column from the sf object
that is then used in the plot method. In this example the st_geometry
finction is used on the `county` polygon in the first plot() call. Then
it used again in the second plot() call on the `foo` point object.
Several other arguments are used in the plot() functions below such
`border = "navy"` sets the outside of the polygon to navy blue, and the
`col = NA` means do not fill in the polygon. Simmilary the `col = "red"`
sets the points in the second plot() call to be red filled. You can look
at the resulting map and guess what `main = "Foo Cases` does in the map.

```{r base-point-map}
plot(st_geometry(county), border = "navyblue", col = NA, main = "Foo Cases")
plot(st_geometry(foo), col = "red3", pch = 16, add = TRUE)
```

A `ggplot2` map is usually better for reports and vignettes as it can be
more readily customized.

```{r ggplot-point-map}

ggplot() +
  geom_sf(data = county, fill = NA, color = "navyblue", linewidth = 0.8) +
  geom_sf(data = foo, color = "red3", size = 1.5) +
  labs(
    title = "Fooflu case locations",
    subtitle = "Synthetic outbreak data"
  ) +
  map_theme
```

You can add more base layers for context. Here, city boundaries are
added behind the points.

```{r city-context-map}
cities <- load_tarrant_spatial("tarrant_cities")

ggplot() +
  geom_sf(data = cities, fill = NA, color = "grey70", linewidth = 0.3) +
  geom_sf(data = county, fill = NA, color = "black", linewidth = 0.8) +
  geom_sf(data = foo, color = "red3", size = 1.5) +
  labs(
    title = "Fooflu case locations with city boundaries",
    subtitle = "Synthetic outbreak data"
  ) +
  map_theme
```

## Static base map tiles

The `base_maps` object contains static map tiles that were previously
retrieved from Google and Stadia Maps. These tiles can be used as a
background for static maps. They are useful when you want more
geographic context than boundary polygons alone can provide.

For example, one available tile is
\`base_maps$stadia$stamen_toner_lite'.

```{r base-map-object}
  plot(base_maps$stadia$stamen_toner_lite, main = "Stamen Toner Lite Map Tile")
```

The map tile can be drawn with the ggmap() function from the ggmap
package.Then the `sf` layers are added on top.

```{r static-basemap-map, message = FALSE, warning = FALSE}

ggmap::ggmap(base_maps$stadia$stamen_toner_lite) +
  geom_sf(data = county, fill = NA, color = "navy", linewidth = 0.8, inherit.aes = FALSE) +
  geom_sf(data = foo, color = "red3", size = 1.4, inherit.aes = FALSE) +
  labs(
    title = "Fooflu case locations",
    subtitle = "Static Stadia/Stamen toner lite base map"
  ) +
  map_theme

```

If your points do not appear where expected, check the CRS first. A
common mistake is to plot longitude/latitude points directly on a base
map without transforming them. All layers should be using the same CRS.

## Coordinate reference systems

A coordinate reference system (CRS) tells `sf` how numeric coordinates
relate to locations on Earth. A CRS identifies more than a unit: it also
describes the datum or reference ellipsoid, coordinate axes, projection,
and area where the system is intended to be used.

### Geographic and projected CRSs

Two broad CRS types are important for this workflow:

1.  **Geographic CRS.** Coordinates are longitude and latitude angles,
    usually expressed in degrees. EPSG:4326 is common for GPS data,
    storage, data exchange, and web services.
2.  **Projected CRS.** Coordinates represent positions on a flat map and
    use linear units such as feet or meters. A projected CRS is designed
    for a particular region and purpose. It is normally the appropriate
    choice for local distance, buffer, area, and density-based cluster
    calculations.

The spatial layers referenced by `geo_sets` are stored in EPSG:4326.
`load_tarrant_spatial()` returns a selected layer in EPSG:4326 by
default, or transforms it when `.crs` is supplied:

```{r load-projected-layer}
county_2276 <- load_tarrant_spatial(
  "tarrant_border",
  .crs = "EPSG:2276"
)
```

Inspect a layer before analysis with:

```{r is-longlat}
sf::st_crs(foo)
sf::st_is_longlat(foo)
sf::st_bbox(foo)
```

Because `foo` is in EPSG:4326, its coordinates are angles rather than
linear distances. One degree of longitude does not represent the same
ground distance everywhere, and degrees should not be treated as feet or
meters in local cluster analysis.

### Assigning a CRS is not transforming a CRS

When creating points from longitude and latitude columns,
`sf::st_as_sf(..., crs = 4326)` assigns the known source CRS. Assigning
a CRS labels the existing coordinate numbers; it does not move them.

`sf::st_transform()` performs a coordinate transformation. It calculates
new coordinate values in the target CRS:

```{r projected-example}
foo_2276 <- sf::st_transform(foo, "EPSG:2276")

sf::st_crs(foo_2276)
sf::st_is_longlat(foo_2276)
```

Do not use `sf::st_set_crs()` or `sf::st_crs(x) <- ...` to make
longitude/latitude coordinates projected. Those operations relabel the
same numbers and can place features in the wrong location. Use them only
to correct missing or incorrect metadata when the true CRS of the
existing coordinates is known.

### CRS requirements for clustering

`scan_cluster()` uses distances between points. It therefore requires:

-   an input `sf` object containing `POINT` geometry;
-   a valid source CRS on the input object; and
-   a projected target CRS with recognized linear units.

The function transforms points to the target `crs`, runs the selected
clustering method in that projected space, and transforms the returned
`clustered` object back to the input CRS. Its default target is
EPSG:2276, NAD83 / Texas North Central (ftUS), which is appropriate for
Tarrant County and uses US survey feet.

The `eps` argument is the neighborhood distance for DBSCAN and the
maximum search distance for OPTICS. Supplying a `units` value makes the
intended distance explicit and allows `scan_cluster()` to convert it to
the target CRS unit:

```{r explicit-distance-units}
one_mile <- units::set_units(1, mile)

foo_dbscan <- scan_cluster(
  points = foo,
  method = "DBSCAN",
  eps = one_mile,
  crs = "EPSG:2276"
)
```

If `eps` is supplied as a bare number, it is interpreted in the linear
unit of the target CRS. For EPSG:2276, `eps = 5280` means 5,280 US
survey feet. A value with units is clearer and is preferred in analysis
code.

### Choosing and checking a CRS

For work in Tarrant County, the `scan_cluster()` default is generally a
reasonable local choice. For another area, select a projected CRS whose
area of use includes the study region and whose distortion properties
fit the analysis. A suitable local State Plane or UTM CRS is usually
better for distance analysis than a worldwide web-map projection such as
EPSG:3857.

Before overlaying or comparing layers, inspect their CRSs. Transform
layers to a common CRS when necessary:

```{r align-layer-crs}
county_for_points <- sf::st_transform(county, sf::st_crs(foo))

identical(sf::st_crs(county_for_points), sf::st_crs(foo))
```

If points appear in the wrong location, distances are implausible, or a
transformation produces empty or extremely large coordinates, check:

-   whether longitude and latitude were supplied in `x, y` order;
-   whether the input CRS describes the existing coordinate values;
-   whether the target CRS covers the study area;
-   whether every mapped layer has compatible CRS information; and
-   whether the distance value and its unit match the epidemiologic
    question.

## Running scan_cluster()

The `scan_cluster()` function identifies clusters in an `sf` POINT
object using one of three density-based methods:

-   DBSCAN,
-   HDBSCAN, or
-   OPTICS.

The input must be an `sf` object with POINT geometry and a valid CRS.

Here is a first HDBSCAN example:

```{r hdbscan-example}
foo_hdbscan <- scan_cluster(
  points = foo,
  method = "HDBSCAN",
  minPts = 7,
  crs = "epsg:2276"  # this is the same as the default
)

foo_hdbscan
```

The returned object is a `clustered` object. It is still an `sf` object,
but it has additional clustering information.

The main added column is `cluster`.

```{r cluster-column}
# First few rows of cases, note the cluster column
foo_hdbscan

# Show the clusters and the number of cases in each
foo_hdbscan$cluster |> 
  table()
```

Cluster `0` means noise. Noise points are points that were not assigned
to a cluster. Positive cluster numbers are detected clusters.

The returned object also stores the original clustering result from the
`hdbscan` function. Use `scan_object()` to retrieve it.

```{r scan-object}
hdbscan_object <- scan_object(foo_hdbscan)

hdbscan_object
```

Use `scan_unit()` to see the distance unit used for the clustering
algorithm.

```{r scan-unit}
scan_unit(foo_hdbscan)
```

## Plotting clustered objects

A `clustered` object has a plot method. The default is to return an
interactive plot.

```{r plot-clustered}
plot(foo_hdbscan)
```

For a static report-style map, you can use `autoplot()`.

```{r autoplot-clustered}
autoplot(foo_hdbscan, title = "Fooflu clusters identified with HDBSCAN", background = base_maps$stadia$stamen_toner_lite)
```

The `autoplot()` method also creates polygons around clusters using
`get_cluster_polys()`. These polygons are helpful for visual
interpretation, but they should not be interpreted as exact exposure
areas.

You can create those polygons directly:

```{r cluster-polys}
foo_polys <- get_cluster_polys(foo_hdbscan, cluster = cluster, ratio = 0.7)

foo_polys
```

Then map the points and polygons together.

```{r cluster-poly-map}
ggplot() +
  geom_sf(data = county, fill = NA, color = "grey50") +
  geom_sf(data = foo_polys, aes(fill = factor(cluster)), alpha = 0.25) +
  geom_sf(data = foo_hdbscan, aes(color = factor(cluster)), size = 1.5) +
  labs(
    title = "Fooflu HDBSCAN clusters",
    fill = "Cluster",
    color = "Cluster"
  ) +
  map_theme
```

## Clustered maps with static base map tiles

If you want the clustered map on top of a static map tile Use the
`background` argument in `autoplot()`.

```{r clustered-basemap-map, message = FALSE, warning = FALSE}


print("TO DO")

#autoplot(
#  foo_hdbscan,
#  background = base_maps$stadia$stamen_toner_lite,
#  title = "Fooflu HDBSCAN clusters",
#  subtitle = "Static base map with clustered points"
#)
```

## Comparing DBSCAN, HDBSCAN, and OPTICS

The three methods available through `scan_cluster()` are related, but
they answer slightly different questions.

| Method | Uses `eps`? | Good for | Practical note |
|:----------------|:------------------:|:----------------|:----------------|
| DBSCAN | Yes | A simple first pass when clusters are expected to have similar density | Results can change substantially when `eps` changes |
| HDBSCAN | Not in the basic call | Clusters with varying density | Often a useful default when you do not know one good distance threshold |
| OPTICS | Yes | Exploring clustering structure across distances | In `scan_cluster()`, OPTICS is followed by `extractXi()` to assign clusters |

The two most important parameters are:

-   `minPts`: the minimum number of nearby points needed to form a dense
    area.
-   `eps`: the search distance used by DBSCAN and OPTICS.

For DBSCAN, `eps` is especially important. In the example below, `eps`
is set to 1.25 miles. Because the clustering CRS is EPSG:2276, the
function converts that value to the CRS unit.

```{r run-dbscan}
foo_dbscan <- scan_cluster(
  points = foo,
  method = "DBSCAN",
  minPts = 7,
  eps = units::set_units(5280 * 1.25, feet),
  crs = "epsg:2276"
)

foo_dbscan
```

HDBSCAN does not require the same single distance threshold in the basic
call.

```{r run-hdbscan}
foo_hdbscan <- scan_cluster(
  points = foo,
  method = "HDBSCAN",
  minPts = 7,
  crs = "epsg:2276"
)

foo_hdbscan
```

OPTICS uses `eps` as a maximum search distance. In `scan_cluster()`, the
OPTICS result is passed through `extractXi()` to create cluster
assignments.

```{r run-optics}
foo_optics <- scan_cluster(
  points = foo,
  method = "OPTICS",
  minPts = 4,
  eps = units::set_units(5280 * 2.55, feet),
  crs = "epsg:2276"
)

foo_optics
```

The summaries can be compared by counting points in each cluster.

```{r compare-cluster-counts}
bind_rows(
  foo_dbscan |>
    st_drop_geometry() |>
    count(cluster, name = "n") |>
    mutate(method = "DBSCAN"),
  foo_hdbscan |>
    st_drop_geometry() |>
    count(cluster, name = "n") |>
    mutate(method = "HDBSCAN"),
  foo_optics |>
    st_drop_geometry() |>
    count(cluster, name = "n") |>
    mutate(method = "OPTICS")
) |>
  select(method, cluster, n) |>
  arrange(method, cluster)
```

And the maps can be compared visually.

```{r plot-dbscan}
print("TO DO")

#autoplot(foo_dbscan, title = "DBSCAN")
```

```{r plot-hdbscan}
# autoplot(foo_hdbscan, title = "HDBSCAN")
```

```{r plot-optics}
#autoplot(foo_optics, title = "OPTICS")
```

The methods may identify different clusters because they define density
differently. This is expected. Cluster detection is exploratory and
should be interpreted with local epidemiologic knowledge.

## Choosing parameters

There is no single correct value for `minPts` or `eps`.

A practical starting approach is:

1.  Choose a projected CRS that is appropriate for the study area.
2.  Choose a distance that has a real-world interpretation.
3.  Choose a minimum point count that reflects the size of cluster you
    care about.
4.  Run the clustering method.
5.  Map the results.
6.  Repeat with other plausible values and compare.

For example, in Tarrant County with `crs = "epsg:2276"`:

```{r compare-eps, eval = FALSE}
scan_cluster(
  points = foo,
  method = "DBSCAN",
  minPts = 7,
  eps = units::set_units(0.5, mile),
  crs = "epsg:2276"
)

scan_cluster(
  points = foo,
  method = "DBSCAN",
  minPts = 7,
  eps = units::set_units(1, mile),
  crs = "epsg:2276"
)

scan_cluster(
  points = foo,
  method = "DBSCAN",
  minPts = 7,
  eps = units::set_units(2, mile),
  crs = "epsg:2276"
)
```

Small `eps` values tend to find only compact clusters. Large `eps`
values may merge nearby groups into larger clusters.

## Important cautions

Spatial clusters are exploratory. They do not prove transmission,
exposure, or a shared source. A cluster can appear because of population
density, geocoding quality, testing patterns, reporting practices, or
the selected clustering parameters.

Always consider:

-   whether the cases are geocoded accurately,
-   whether duplicate addresses or institutional addresses are present,
-   whether the CRS is appropriate,
-   whether the selected distance threshold makes sense,
-   whether population distribution should be considered, and
-   whether the result matches field investigation knowledge.

## Complete example

This is a compact version of the workflow.

```{r complete-example, eval = FALSE}
library(tarr)
library(sf)
library(dplyr)
library(ggplot2)
library(units)

county <- load_tarrant_spatial("tarrant_border")

foo <- synthetic_outbreak |>
  filter(condition == "fooflu") |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

foo_clusters <- scan_cluster(
  points = foo,
  method = "HDBSCAN",
  minPts = 7,
  crs = "epsg:2276"
)

#autoplot(foo_clusters, title = "Fooflu HDBSCAN clusters")
```

For a static base map tile:

```{r complete-basemap-example, eval = FALSE}

plot(foo_clusters)

```
