---
title: "Maps and Clusters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Maps and Clusters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  fig.width = 8,
  fig.height = 6,
  out.width = "100%"
)
```

## Overview

This vignette introduces a practical mapping and clustering workflow using `tarr`, `sf`, and the density-based clustering methods available through `scan_cluster()`.

The examples use the `synthetic_outbreak` data frame included with the `tarr` package. It  contains simulated cases of
two fake conditions in Tarrant County. The same workflow can be adapted to a real line list that has one row per case
and columns for longitude and latitude.

By the end of this vignette, you should be able to:

- load common Tarrant County spatial layers,
- convert a line list into an `sf` POINT object,
- plot case points on simple static maps,
- use package-provided static base map tiles,
- understand why coordinate reference systems matter,
- choose a projected coordinate reference system for clustering,
- run `scan_cluster()`, and
- compare DBSCAN, HDBSCAN, and OPTICS results.

This vignette assumes that you are fairly new to GIS work in R. The examples are intentionally verbose so that you can see each step.

## Packages used

```{r packages, message = FALSE, warning = FALSE}
library(tarr)
library(maptiles)
library(sf)
library(dplyr)
library(ggplot2)
library(units)
library(dbscan)
```

The most important package for GIS work is `sf`. An `sf` object is a data frame with a special geometry column. The
regular columns contain attributes such as condition, sex, onset date, or case ID. The geometry column contains the
spatial information, such as a point location or polygon boundary.

## Tarrant County spatial layers

The `tarr` package includes several commonly used Tarrant County spatial layers. The named list `geo_sets` stores the
file paths for these layers. The helper function `load_tarrant_spatial()` loads one of these layers as an `sf` object.

Common layers include:

| Name | Description | Geometry |
|:---|:---|:---|
| `ZCTA` | ZIP Code Tabulation Areas | Polygon |
| `tarrant_border` | Tarrant County boundary | Polygon |
| `tarrant_tracts` | Census tracts in Tarrant County | Polygon |
| `tarrant_blocks` | Census blocks in Tarrant County | Polygon |
| `tarrant_cities` | Boundaries of cities that are partially or wholly in Tarrant County | Polygon |
| `surrounding_borders` | Borders of counties surrounding Tarrant County | Polygon |
| `isd` | Independent school districts | Polygon |

You can see the available layer names with:

```{r geo-set-names}
names(geo_sets)
```

The usual way to load one layer is:

```{r load-county}
county <- load_tarrant_spatial("tarrant_border")

county
```

The result is an `sf` object. You can check the geometry type and coordinate reference system:

```{r inspect-county}
sf::st_geometry_type(county)
sf::st_crs(county)
```

The default coordinate reference system for these layers is EPSG:4326, which stores coordinates as longitude and latitude.
You can also request a different coordinate reference system when loading a layer. This is useful when you want a layer to match another spatial object.

```{r load-county-web}
county_3857 <- load_tarrant_spatial("tarrant_border", .crs = 3857)

sf::st_crs(county_3857)
```

## A first static map

Before mapping cases, it helps to map the county boundary and nearby counties.

```{r simple-boundary-map}
surrounding <- load_tarrant_spatial("surrounding_borders")

plot(st_geometry(surrounding), border = "grey60")
plot(st_geometry(county), border = "red3", lwd = 2, add = TRUE)
```

This uses base R plotting. It is quick and useful for checking that a spatial layer loaded correctly.
The same map can be made with `ggplot2`:

```{r ggplot-boundary-map}
ggplot() +
  geom_sf(data = surrounding, fill = NA, color = "grey70") +
  geom_sf(data = county, fill = NA, color = "red3", linewidth = 0.8) +
  labs(title = "Tarrant County and surrounding counties") +
  map_theme
```

The `map_theme` object is a ggplot2 theme provided by `tarr`. It removes axis labels and grid lines that are usually not helpful in 
mapping local areas and sets font characterists. .

## Making a POINT sf object from a line list

A line list is usually stored as a data frame with one row per case and attributes for each case in columns
(e.g., age, address, etc.). If the line list has coordinates like longitude/latitude columns, it can be converted into an `sf` POINT
object.

The package `synthetic_outbreak` data frame  has columns named `lon` and `lat`.

```{r look-at-linelist}
synthetic_outbreak |>
  select(condition, onset, lon, lat) |>
  head()
```

For this vignette, we will focus on the fake condition `fooflu`.

```{r filter-fooflu}
foo_cases <- synthetic_outbreak |>
  filter(condition == "fooflu")

nrow(foo_cases)
```

Now convert the ordinary data frame foo_cases into an `sf` object that will be called `foo`.

```{r line-list-to-sf}
foo <- foo_cases |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

foo
```

The `coords` argument identifies the coordinate columns. Longitude is the x-coordinate and latitude is the y-coordinate,
so the order is `c("lon", "lat")`.

The `crs = 4326` argument tells `sf` that these coordinates are longitude and latitude coordinates. Coordinate systems
can be expressed in decimal degrees, feet, yards, meters, etc.  Coordinate reference systems (CRS) give a model of how the 
earth is modeled.  Some CRS project an area of the pearth as a plane, perhaps in meters or feet and can be valid over 
one area om the earth. Other CRS can ue world coordinates such as decimal degrees often expressed as longitude/latitude
or lon/lat.   Having a CRS set for an sf object is important. If it is an `sf` object does not have the correct CRS, GIS
software including R cannot reliably transform it, measure distances, or overlay it with other layers.

You can confirm that `foo` is now an `sf` POINT object and also see what the crs is for the object. The print out for 
crs of foo is quite long.  Note how it states the ANGLEUNIT is "degree", the area it covers `World`.  A special value 
at the bottom is `ID[ESPG, 4326]. All of the map tiles, and spatial data sets in this package use ESPG:4326.  This 
the most commonly used refeence system when web mapping tools are used.

```{r inspect-foo}
sf::st_geometry_type(foo)
sf::st_crs(foo)
```

## Plotting case points

A quick base R plot is often the easiest first check.  Note the st_geomtery() function.  It pulls the geometry column from
the sf object that is then used in the plot method.  In this example the st_geometry finction is used on the `county` polygon 
in the first plot() call.  Then it used again in the second plot() call on the `foo` point object. Several other
arguments are used in the plot() functions below such  `border = "navy"` sets the outside of the polygon to navy blue,
and the `col = NA` means do not fill in the polygon.  Simmilary the `col = "red"` sets the points in the second plot()
call to be red filled.  You can look at the resulting map and guess what `main = "Foo Cases` does in the map. 

```{r base-point-map}
plot(st_geometry(county), border = "navyblue", col = NA, main = "Foo Cases")
plot(st_geometry(foo), col = "red3", pch = 16, add = TRUE)
```

A `ggplot2` map is usually better for reports and vignettes as it can be more readily customized.

```{r ggplot-point-map}

ggplot() +
  geom_sf(data = county, fill = NA, color = "navyblue", linewidth = 0.8) +
  geom_sf(data = foo, color = "red3", size = 1.5) +
  labs(
    title = "Fooflu case locations",
    subtitle = "Synthetic outbreak data"
  ) +
  map_theme
```

You can add more base layers for context. Here, city boundaries are added behind the points.

```{r city-context-map}
cities <- load_tarrant_spatial("tarrant_cities")

ggplot() +
  geom_sf(data = cities, fill = NA, color = "grey70", linewidth = 0.3) +
  geom_sf(data = county, fill = NA, color = "black", linewidth = 0.8) +
  geom_sf(data = foo, color = "red3", size = 1.5) +
  labs(
    title = "Fooflu case locations with city boundaries",
    subtitle = "Synthetic outbreak data"
  ) +
  map_theme
```

## Static base map tiles

The `base_maps` object contains static map tiles that were previously retrieved from Google and Stadia Maps. These tiles can be used as a background for static maps. They are useful when you want more geographic context than boundary polygons alone can provide.

For example, one available tile is `base_maps$stadia$stamen_toner_lite'.  

```{r base-map-object}
  plot(base_maps$stadia$stamen_toner_lite, main = "Stamen Toner Lite Map Tile")
```


The map tile can be drawn with the ggmap() function from the ggmap package.Then the `sf` layers are added on top.

```{r static-basemap-map, message = FALSE, warning = FALSE}

ggmap::ggmap(base_maps$stadia$stamen_toner_lite) +
  geom_sf(data = county, fill = NA, color = "navy", linewidth = 0.8, inherit.aes = FALSE) +
  geom_sf(data = foo, color = "red3", size = 1.4, inherit.aes = FALSE) +
  labs(
    title = "Fooflu case locations",
    subtitle = "Static Stadia/Stamen toner lite base map"
  ) +
  map_theme

```

If your points do not appear where expected, check the CRS first. A common mistake is to plot longitude/latitude points directly on a base map without transforming them.  All layers should be using the same CRS.  

## Coordinate reference systems

A coordinate reference system, or CRS, tells R how coordinates relate to places on Earth.

There are two broad CRS types that matter for this vignette:

1. Geographic CRS  
   Coordinates are stored as longitude and latitude. EPSG:4326 is a common example. This is convenient for storage, sharing, and many web mapping tasks.

2. Projected CRS  
   Coordinates are stored on a flat projected surface. Units are usually feet or meters. Projected coordinate systems are usually better for distance calculations.

You can check whether an object is longitude/latitude with:

```{r is-longlat}
st_is_longlat(foo)
```

Because `foo` is in EPSG:4326, its coordinates are degrees. Degrees are not a good unit for density-based clustering. A difference of one degree of longitude is not the same physical distance everywhere on Earth.

The `scan_cluster()` function performs distance-based clustering. For that reason, it needs a projected CRS for clustering. Its default clustering CRS is `"epsg:2276"`, which is appropriate for North Central Texas and uses feet. This makes values like `5280` meaningful as about one mile.

```{r projected-example}
foo_2276 <- st_transform(foo, "epsg:2276")

st_crs(foo_2276)
st_is_longlat(foo_2276)
```

For work in Tarrant County, the scan_cluster() default `crs = "epsg:2276"` is usually a good choice. For work somewhere else, choose a local projected CRS or an appropriate UTM CRS for the study area.

## Running scan_cluster()

The `scan_cluster()` function identifies clusters in an `sf` POINT object using one of three density-based methods:

- DBSCAN,
- HDBSCAN, or
- OPTICS.

The input must be an `sf` object with POINT geometry and a valid CRS.

Here is a first HDBSCAN example:

```{r hdbscan-example}
foo_hdbscan <- scan_cluster(
  points = foo,
  method = "HDBSCAN",
  minPts = 7,
  crs = "epsg:2276"  # this is the same as the default
)

foo_hdbscan
```

The returned object is a `clustered` object. It is still an `sf` object, but it has additional clustering information.

The main added column is `cluster`.

```{r cluster-column}
# First few rows of cases, note the cluster column
foo_hdbscan

# Show the clusters and the number of cases in each
foo_hdbscan$cluster |> 
  table()
```

Cluster `0` means noise. Noise points are points that were not assigned to a cluster. Positive cluster numbers are detected clusters.

The returned object also stores the original clustering result from the `hdbscan` function. Use `scan_object()` to retrieve it.

```{r scan-object}
hdbscan_object <- scan_object(foo_hdbscan)

hdbscan_object
```

Use `scan_unit()` to see the distance unit used for the clustering algorithm.

```{r scan-unit}
scan_unit(foo_hdbscan)
```

## Plotting clustered objects

A `clustered` object has a plot method. The default is to return an interactive plot.

```{r plot-clustered}
plot(foo_hdbscan)
```

For a static report-style map, you can use `autoplot()`.

```{r autoplot-clustered}
autoplot(foo_hdbscan, title = "Fooflu clusters identified with HDBSCAN", background = base_maps$stadia$stamen_toner_lite)
```

The `autoplot()` method also creates polygons around clusters using `get_cluster_polys()`. These polygons are helpful for visual interpretation, but they should not be interpreted as exact exposure areas.

You can create those polygons directly:

```{r cluster-polys}
foo_polys <- get_cluster_polys(foo_hdbscan, cluster = cluster, ratio = 0.7)

foo_polys
```

Then map the points and polygons together.

```{r cluster-poly-map}
ggplot() +
  geom_sf(data = county, fill = NA, color = "grey50") +
  geom_sf(data = foo_polys, aes(fill = factor(cluster)), alpha = 0.25) +
  geom_sf(data = foo_hdbscan, aes(color = factor(cluster)), size = 1.5) +
  labs(
    title = "Fooflu HDBSCAN clusters",
    fill = "Cluster",
    color = "Cluster"
  ) +
  map_theme
```

## Clustered maps with static base map tiles

If you want the clustered map on top of a static map tile
Use the `background` argument in `autoplot()`.

```{r clustered-basemap-map, message = FALSE, warning = FALSE}


print("TO DO")

#autoplot(
#  foo_hdbscan,
#  background = base_maps$stadia$stamen_toner_lite,
#  title = "Fooflu HDBSCAN clusters",
#  subtitle = "Static base map with clustered points"
#)
```

## Comparing DBSCAN, HDBSCAN, and OPTICS

The three methods available through `scan_cluster()` are related, but they answer slightly different questions.

| Method | Uses `eps`? | Good for | Practical note |
|:---|:---:|:---|:---|
| DBSCAN | Yes | A simple first pass when clusters are expected to have similar density | Results can change substantially when `eps` changes |
| HDBSCAN | Not in the basic call | Clusters with varying density | Often a useful default when you do not know one good distance threshold |
| OPTICS | Yes | Exploring clustering structure across distances | In `scan_cluster()`, OPTICS is followed by `extractXi()` to assign clusters |

The two most important parameters are:

- `minPts`: the minimum number of nearby points needed to form a dense area.
- `eps`: the search distance used by DBSCAN and OPTICS.

For DBSCAN, `eps` is especially important. In the example below, `eps` is set to 1.25 miles. Because the clustering CRS is EPSG:2276, the function converts that value to the CRS unit.

```{r run-dbscan}
foo_dbscan <- scan_cluster(
  points = foo,
  method = "DBSCAN",
  minPts = 7,
  eps = units::set_units(5280 * 1.25, feet),
  crs = "epsg:2276"
)

foo_dbscan
```

HDBSCAN does not require the same single distance threshold in the basic call.

```{r run-hdbscan}
foo_hdbscan <- scan_cluster(
  points = foo,
  method = "HDBSCAN",
  minPts = 7,
  crs = "epsg:2276"
)

foo_hdbscan
```

OPTICS uses `eps` as a maximum search distance. In `scan_cluster()`, the OPTICS result is passed through `extractXi()` to create cluster assignments.

```{r run-optics}
foo_optics <- scan_cluster(
  points = foo,
  method = "OPTICS",
  minPts = 4,
  eps = units::set_units(5280 * 2.55, feet),
  crs = "epsg:2276"
)

foo_optics
```

The summaries can be compared by counting points in each cluster.

```{r compare-cluster-counts}
bind_rows(
  foo_dbscan |>
    st_drop_geometry() |>
    count(cluster, name = "n") |>
    mutate(method = "DBSCAN"),
  foo_hdbscan |>
    st_drop_geometry() |>
    count(cluster, name = "n") |>
    mutate(method = "HDBSCAN"),
  foo_optics |>
    st_drop_geometry() |>
    count(cluster, name = "n") |>
    mutate(method = "OPTICS")
) |>
  select(method, cluster, n) |>
  arrange(method, cluster)
```

And the maps can be compared visually.

```{r plot-dbscan}
print("TO DO")

#autoplot(foo_dbscan, title = "DBSCAN")
```

```{r plot-hdbscan}
# autoplot(foo_hdbscan, title = "HDBSCAN")
```

```{r plot-optics}
#autoplot(foo_optics, title = "OPTICS")
```

The methods may identify different clusters because they define density differently. This is expected. Cluster detection is exploratory and should be interpreted with local epidemiologic knowledge.

## Choosing parameters

There is no single correct value for `minPts` or `eps`.

A practical starting approach is:

1. Choose a projected CRS that is appropriate for the study area.
2. Choose a distance that has a real-world interpretation.
3. Choose a minimum point count that reflects the size of cluster you care about.
4. Run the clustering method.
5. Map the results.
6. Repeat with other plausible values and compare.

For example, in Tarrant County with `crs = "epsg:2276"`:

```{r compare-eps, eval = FALSE}
scan_cluster(
  points = foo,
  method = "DBSCAN",
  minPts = 7,
  eps = units::set_units(0.5, mile),
  crs = "epsg:2276"
)

scan_cluster(
  points = foo,
  method = "DBSCAN",
  minPts = 7,
  eps = units::set_units(1, mile),
  crs = "epsg:2276"
)

scan_cluster(
  points = foo,
  method = "DBSCAN",
  minPts = 7,
  eps = units::set_units(2, mile),
  crs = "epsg:2276"
)
```

Small `eps` values tend to find only compact clusters. Large `eps` values may merge nearby groups into larger clusters.

## Important cautions

Spatial clusters are exploratory. They do not prove transmission, exposure, or a shared source. A cluster can appear because of population density, geocoding quality, testing patterns, reporting practices, or the selected clustering parameters.

Always consider:

- whether the cases are geocoded accurately,
- whether duplicate addresses or institutional addresses are present,
- whether the CRS is appropriate,
- whether the selected distance threshold makes sense,
- whether population distribution should be considered, and
- whether the result matches field investigation knowledge.

## Complete example

This is a compact version of the workflow.

```{r complete-example, eval = FALSE}
library(tarr)
library(sf)
library(dplyr)
library(ggplot2)
library(units)

county <- load_tarrant_spatial("tarrant_border")

foo <- synthetic_outbreak |>
  filter(condition == "fooflu") |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

foo_clusters <- scan_cluster(
  points = foo,
  method = "HDBSCAN",
  minPts = 7,
  crs = "epsg:2276"
)

#autoplot(foo_clusters, title = "Fooflu HDBSCAN clusters")
```

For a static base map tile:

```{r complete-basemap-example, eval = FALSE}

plot(foo_clusters)

```
