---
title: "Packge Overview"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Packge Overview}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}

library(tidyverse, quietly = TRUE)
library(tarr, quietly      = TRUE)

# Setup for examples (feel free to change these defaults)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE,
  echo = TRUE,
  eval = FALSE)


```

## Overview & goals

The tarr package includes various functions and lists to assist
Epidemiologists at the Tarrant County Public Health Department. It was
originally written to support disease data manipulation and creating
charts. Over time additional functions, data sets, and list variables have been
created to simplify workflows and add capabilities.

This vignette reviews installation, standard folders accessed, helpful
lists, and helper functions. Addtional vignettes introduce important package
capabilities including:

**Reading and Processing** NBS and EpiTrax data exports. Covers the
functions to read the data and how the master file of all case records
is constructed and maintained.

**Map & cluster** describes the geographic layers provided, functions
that aid mapping, and spatial cluster detection.

## Installation & project setup

The package is locally maintained and as of
`r format(Sys.Date(), "%B %d, %Y")`, may be found on the Epidemiology
shared drive folder as `r file.path("G:/Data/Packages", paste0("tarr.", utils::packageVersion("tarr")))`.
 Several other packages are needed for tarr to operate properly. Since
the package is not hosted on CRAN, dependencies are not automatically
installed. Source the `tarr_depend.r` in the same foler as the package. This will install
most, if not all, of the R packages needed for the tarr package to
properly function. Two related packages are also available for installation:

 - tarr.pop for population data and manipulation. 
 - tarr.brand for Tarrant County branding colors and seals

#### Note

Use of the tarr package is expected to occur in the secure environment of the
county network, i.e., behind the county firewall. Data from NEDSS and
EpiTrax contain personal health information. Outside of the network
You can practice using some of the functions outside of the county network by 
using the `synthetic_outbreak` tibble available within the package. 

## Standard folders

The tarr package reads and processes data placed in standard folders or
sub-folders on the shared epidemiology (E:) and data (G:) drives. The
different folder paths are kept in a list variable called **paths**
which references the main folders. The table below shows the "path" to
each folder referenced. In turn each of those folders have sub-folders
with documents and data.

| Name | Folder | Data Types |
|:---------------|:----------------------------|:--------------------------|
| data | G:/Data/Commmunicable Disease/Current Year | Exported csv files from EpiTrax and NBS |
| communicable | G:/Data/Commmunicable Disease | Parent folder for "data" folder |
| scripts | G:/Data/Commmunicable Disease/R Scripts | A folder of R scripts to work on exported data |
| cities | G:/Data/GIS/Shape Files/Tarrant/Cities | GIS polygon file for Tarrant Cities |
| spatial | G:/Data/GIS/Shape Files | General GIS files with several subfolders |
| population | G:/Data/Population | Census, projections, and estimates files |
| monthReportCode | E:/Monthly Report/Code | The folder for the monthly report R project |
| monthReport | E:/Monthly Report | Subfolders for each month's report |
| arbovirus | E:/Arboviral Diseases | Arboviral disease data files and guidance |
| immunization | E:/Immunization | Immunization data |
| flu | PublicHealth/Shared/EPI/Influenza | Influenza surveillance folder |

Here is an example of using the `paths` variable to list all the files and folders in the 
in the spatial folder:

```{r paths, eval=TRUE}
# The path to the data folder
paths$spatial

# List the file names in the spatial  folder
files <- list.files(path = paths$spatial)  # Search for all files
head(files)  # View the first six in the folder
rm(files)
```

## Miscellaneous helpers

Several utility functions are included. Among these are functions to
retrieve MMWR weeks and dates, read and write to the clipboard, some
character string manipulations, detecting and converting character
vectors to logical vectors.

#### MMWR Weeks

A set of functions are provided for retrieving MMWR weeks (a.k.a. epi
weeks) and dates, The basic `mmwrWeek(date)` function returns the
numeric MMWR week of a date vector. Conversely you can get the first date 
from an MMWR week and year using `mmwrWeekBegin(weekNum, yr)`,
the last date of the week `mmwrWeekEnd(weeknum, yr)`, and the month of the
MMWR week passing he week number and year 
`mmwrWeekMonth(weeknum, yr, label, abbr)`.

You may use either a `Date` or a `POSIXct datetime` to retrieve the MMWR
week number. This is useful if you want to calculate the MMWR week for
dates like onset, birth, report, diagnosis, etc. Here's an example of
using onset and getting the week number.

```{r mmwr_week, eval=TRUE}

# sample data
data("synthetic_outbreak", package = "tarr")
foo_flu <- synthetic_outbreak |> 
           dplyr::filter(condition == "fooflu") 

# add the MMWR Week number when onset occurred
foo_flu$mmwr_week <- mmwrWeek(foo_flu$onset)

# look at onset dates and epi week for the first few records
foo_flu[, c("onset", "mmwr_week")] |> head()

# table by week
janitor::tabyl(foo_flu$mmwr_week)
```

You may wish to convert an MMWR week to an actual date in a given year.
Two functions can do that; mmwrWeekBegin() returns the date on the
Sunday of the week and mmwrWeekEnd() gives the Saturday of that week. If
you prefer the mid week date,it's as simple as giving a week number and
year to mmwrWeekBegin() + 3, to get the date on Wednesday.

In addition to the week dates, the mmwrWeekMonth() function can be used
to get the month number, full name, or abbreviation. By default it
returns the month number for a given weekNum and yr. When the `label`
argument is set to TRUE, the function returns the full month name. When
the `label` and `abbrv` arguments are both set to TRUE, the month
abbreviation is returned.

Using the date instead of the week number can help in interpreting
charts as a person does not have to mentally calculate the month when an
event occurs. The ability to convert a week number to a date in a
specific year also allows you to show different years data using the
same date scale.

```{r mmwr_begin_end, eval=TRUE}
# show dates and month information from mmwrWeek
foo_flu <- foo_flu |> 
  head() |> 
  mutate(begin_date = mmwrWeekBegin(weekNum = mmwr_week, yr = 2020),
         end_date   = mmwrWeekEnd(weekNum = mmwr_week, yr = 2020),
         month_num  = mmwrWeekMonth(mmwr_week, 2020),
         month_name = mmwrWeekMonth(mmwr_week, 2020, label = TRUE)
         )  
  foo_flu[, c("mmwr_week", "begin_date", "end_date", "month_num", "month_name")] |> head()
```

#### Read and write to the clipboard

Two functions are provided to interact with the clipboard; `read.clip()`
and `write.clip(x)`. Both act on tabular data like a data frame and are
simple wrappers around the clipr package functions read_clip_tbl() and
write_clip(). Passing a data frame to write.clip() will place a copy of
the data frame on the clipboard which can then be posted in another
program like Excel or Word. Conversely if you wish to move tabular data
from a program like Excel, highlight the data including headers, and
press 'ctrl-c' to copy the data to the clipboard. In yor R script you
can assign the clipboard data to a data frame like so `df <- read.clip()`.

#### Special string functions

Several simple character string functions are provided to extract
classes of characters. They are collectivley named "mash" or "only"
indicating the functions return only certain character types and can
result in the character string being "mashed". They are useful when
you want to have only alphanumeric characters in a string to improve
string matching.

The functions `only_digit(string)` and `mash_phone(string)` are
synonymous, returning only digit characters.

```{r mash_phone, eval=TRUE}

# phone numbers formatted differently
phone_numbers <- c("817/321-4700", "817-321-2366", "(214)972-9840")

# show only the digits
mash_phone(phone_numbers)

# or using the synonmous function
only_digits(phone_numbers) 
rm(phone_numbers)
```

The `only_alpha()` and `mash_name()` functions are also synonymous,
returning only alphabetic characters.

```{r only_alpha, eval=TRUE}
addr_roads <- c("5001 Riverside", "1101 South Main Street", "Interstate 35", "Highway 287", "Hwy 10", "200 Weatherford Street")
  
# only alphabetic characters
mash_name(addr_roads)
only_alpha(addr_roads)

rm(addr_roads)
```

#### Character to logical vectors

In many data sets, character strings containing Yes/No or integers 0/1 are used for a TRUE/FALSE variable. Converting
these vectors to an R logical vector can be advantageous to simplify analysis. The "Yes/No" set of functions are used
for this:

-   `is_yn(vec)` checks if vector is a "Yes/No" or 0/1 variable. Returns TRUE
    if all the values fall in a YES/NO/Unknown value set regardless of
    character case. Those sets include "Yes/No/Unknown" or Unk, "Y/N/U",
    or for numeric vectors 0 (FALSE), 1 (TRUE), NA (missing). If the vector is
    "Yes/No" it returns TRUE. An attribute of the returned object called "values" is set with
    the "Yes/No/Unknown" value set used in the vector.

-   `id_yn_cols(df)`identifies the "Yes/No" columns in a data frame. The
    function returns the column names that can be converted to a logical
    vector.

-   `yn_2_logical(vec)` takes a "Yes/No" vector and converts it to a
    logical vector. "Unknown" values are converted to NA.

```{r yn_funcs, eval=TRUE}

recovered <- sample(x = c("Yes", "No", "Unk"), size = 15, replace = TRUE, prob = c(0.50, 0.30, 0.20))
print(recovered)

is_yn(recovered)

# the original and converted vectors together in a data frame
rec_lgl <- yn_2_logical(recovered)
rec_df <- data.frame("recover_string" = recovered, "recover_logical" = rec_lgl)

# show the data frame 
rec_df

rm(recovered, rec_lgl, rec_df)

```

#### General vector conversion function

It is not unusual to have two vector of two different types, but you
want them to be the same type. R and used package will have cpnversion functions
There are many conversion 'as' functions available in R,like;
as.numeric(), as.character(), as.logical to convert vectors to a
different type. The `convert_type(vec1, vec2)` is a generalization of
this concept using a template vector. The function attempts to convert
vec1 to template vector passed as vec2 . It works well for base types
where conversion in R is allowed.

```{r type_conversion, eval=TRUE}

# Dates represented as character string all in year-month-day format
char_dates <- c("2022-05-15", "2016/12/25", "1983-06-25")
date <- as.Date("1990-04-24")

converted <- convert_type(vec1 = char_dates, vec2 = date)
converted
class(converted)
```
