---
title: "The Tarr Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Tarr Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```r
# Setup for examples (feel free to change these defaults)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = #>,
  fig.retina = 2,
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE,
  echo = TRUE,
  eval = FALSE
)
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## 1. Overview & goals The tarr package includes various functions and variables to assist Epidemiologists at the
Tarrant County Public Health Department.  It was originally written to support disease data manipulation and creating
charts.  Over time the functions, data sets, and list variables have been updated to simplify workflows and add 
capabilities.

Separate vignettes introduce different topics including:

**Reading and Processing** NBS and EpiTrax data exports. This document covers the functions to read the data and how 
master the master file of all case records is constructed and maintained.  Combining 
level 
**Ages and Age Groups** demonstrates the functions for calculating ges and age groups. I also covers the class "age_group"
and how it can be used.
**Map & cluster** describes the GIS layers provided, functions that aid mapping, and spatial cluster detection

This vignette reviews installation, standard folders accessed, helpful variables provided, and useful helper functions.

## 2. Installation & project setup
The package is locally maintained and as of `r format(Sys.Date(), "%B %d, %Y")`, may be found on the Epidemiology shared
drive in the "G:/Data/Packages/" folder as "tarr_0.0.0.xxxx.zip.  The xxxx is actually  number that changes as the
package is updated. Several other packages are needed for tarr to operate properly. Since the package is not hosted on
CRAN, dependencies are not automatically installed.  To address this, run the tarr_depend.r in RStudio.  This will
install most, if not all, of the R packages needed for the tarr package to properly function.  In addition it also
installs tarr.population for population figures and tarr.brand that makes Tarrant County branding colors and items
available.

```{r setup, echo=FALSE, warnings=FALSE, include=FALSE}
# install.packages(tarr)                # if released on CRAN
# remotes::install_github(<org>/<repo>) # dev version for future use. 
library(tidyverse, quietly = TRUE)
library(tarr, quietly      = TRUE)
```

Use of the package is expected to occur behind the county firewall in a secure environment, as the data used will 
contain personal health information. You can practice using some of the functions outside of the county network
by using the synthetic_outbreak data frame available within the package. 

## 3. Standard folders
The tarr package reads and processes data placed in standard folders or sub- folderson the Epidemiology shared drive (E:\)
and shared data drive (G:\). The different folder paths are kept in a list variable called **paths** which references 
several folders.  The table below shows the "path" to each folder referenced.  In turn each of those folders have 
sub-folders with documents and data.



|Name            |Folder                                      |Data Types                                |
|:---------------|:-------------------------------------------|:-----------------------------------------|
|data            |G:/Data/Commmunicable Disease/Current Year  |Exported csv files from EpiTrax and NBS   |
|communicable    |G:/Data/Commmunicable Disease               |Parent folder for "data" folder           |
|scripts         |G:/Data/Commmunicable Disease/R Scripts     |A folder of R scripts to work on exported data|
|cities          |G:/Data/GIS/Shape Files/Tarrant/Cities      |GIS files for Tarrant Cities              |
|spatial         |G:/Data/GIS/Shape Files                     |General GIS files with several subfolders |
|population      |G:/Data/Population                          |Saved census/projections/estimates files  | 
|monthReportCode |E:/Monthly Report/Code                      |The folder for monthly report R project   |
|monthReport     |E:/Monthly Report                           |Subfolders of each months report          |
|arbovirus       |E:/Arboviral Diseases                       |Arboviral disease data files and guidance |
|immunization    |E:/Immunization                             |Immunization data                         |
|flu             |PublicHealth/Shared/EPI/Influenza           |Influenza surveillance folder             |


An example of using the paths variable. Say you want to list all the *.csv files in the data folder:

```{r}
# The path to the data older
paths$data

# Get the csv files in the older
files <- list.files(path = paths$data, pattern = "\\.csv$")  # Search for csv as the file extension
head(files)  # View the first six that were found
rm(files)
```

---
## 4. Miscellaneous helpers

Common utility functions tha are useful in other contexts and across epi workflows.

The mmwr_* functions are helpers to retreive weeks or dates when working with MMWR weeks also know as epi weeks.
Given a date, such as an onset or other event date, the mmwrWeek() function returns the week number.  

```r

# sample data
foo_flu <- filter(synthetic_outbreak, condition == "fooflu")

# add a MMWR Week number 
fooflu$mmwr_week <- mmwrWeek(fooflu$onset)

# look at the first few records of those columns 
fooflu$onset, mmwr_week) |> head()
```


You may wish to convert an MMWR week to an actual date in a given year.  Two functions can do that; mmwrWeekBegin() returns
the date on the Sunday of the wee and mmwrWeekEnd() gives the Satruday of that week.  If you prefer the mid week dat,it's
simple as giving a week number and year t mmwrWeekBegin() + 3, toget the Wednesday date. 

```r mmwr_begin_end
foo_flu |> 
  head() |> 
  mutate(begin_date = mmwrWeekBegin(weekNum = mmwrWeek, yr = 2020)) |> 
  pull(begin_date)

```

----
## 6. Population denominators

Population pulls are provided by the companion package **tarr.population**.

```r
# Example (adapt to your dims)
# pop <- tarr.population::population(
#   year = 2023,
#   area.name = Tarrant County,
#   sex = c(Female,Male),
#   age.char = levels(nbs$age_char),
#   race = All, ethnicity = All
# )

# Compute rates after joining cases to denominators
# rates <- dplyr::left_join(case_counts, pop, by = c(year,area.name,sex,age.char)) |>
#   dplyr::mutate(rate = 1e5 * cases / population)
```

## 8. End‑to‑end mini example

```r
nbs <- nbs_import(data/nbs/) |> nbs_process()
nbs$age       <- age_calc(nbs$birth_date, nbs$onset_date)
nbs$age_group <- as.age_group(age_cat(nbs$age))

cases_sf <- sf::st_as_sf(nbs, coords = c(lon,lat), crs = 4326)
cl <- scan_cluster(cases_sf, method = hdbscan, minPts = 5)

autoplot(cl) + map_theme
```