---
title: "Package Overview"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Package Overview}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}

library(tarr, quietly = TRUE)

# Setup for examples (feel free to change these defaults)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 2,
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE,
  echo = TRUE,
  eval = FALSE)


```

## Overview and goals

The `tarr` package includes functions, data sets, and lists that support
epidemiologists at Tarrant County Public Health. It was originally
written to support disease-data manipulation and chart creation. Over
time, additional functions and data have been added to simplify local
workflows and extend the package's capabilities.

This vignette reviews installation, standard folders, useful lists, and
helper functions. Additional vignettes introduce important
package capabilities, including:

- **Reading and processing** covers NBS and EpiTrax data exports,
  including functions that read the data and maintain the master file of
  case records. See
  `vignette("epitrax-and-nbs-workflow", package = "tarr")`.

- **Maps and clusters** describes the provided geographic layers,
  mapping helpers, and spatial-cluster detection. See
  `vignette("maps-and-clusters", package = "tarr")`.

## Installation & project setup

The package is intended to be installed on Tarrant County Windows
workstations. It is maintained locally and distributed from
`G:/Data/Packages` as a Windows binary ZIP file. The filename includes
the package version, for example `tarr_<version>.zip`.

Because `tarr` is not hosted on CRAN, install its dependencies before
installing the package. Source `tarr_depend.r` from the same shared-drive
folder:

```{r install-dependencies, eval=FALSE}
source("G:/Data/Packages/tarr_depend.r")
```

### Install the Windows binary with RStudio

1. Open the Windows version of RStudio.
2. Source `G:/Data/Packages/tarr_depend.r` to install required package
   dependencies.
3. Select **Tools > Install Packages**.
4. For **Install from**, select
   **Package Archive File (.zip; .tar.gz)**.
5. Select the desired R library under **Install to Library**.
6. Browse to `G:/Data/Packages` and select the current `tarr` binary ZIP
   file.
7. Click **Install** and restart the R session after installation.

Verify the installation from the RStudio console:

```{r verify-installation, eval=FALSE}
library(tarr)
packageVersion("tarr")
```

Two related packages are also available for separate installation:

- `tarr.pop` provides population data and population-data helpers.
- `tarr.brand` provides Tarrant County branding colors and seals.

### Protected health information

> **Warning:** NEDSS and EpiTrax exports may contain protected health
> information (PHI). Use real case data only within the approved secure
> county environment and behind the county firewall. Do not copy, upload,
> email, or process identifiable case data on personal devices, public
> services, or systems that have not been approved for PHI. Follow all
> applicable Tarrant County privacy, security, retention, and minimum-
> necessary-use requirements.

Outside the secure county environment, use only synthetic or properly
de-identified data. The bundled `synthetic_outbreak` tibble is available
for practice and examples.

## Standard folders

The `tarr` package is intended for local use with the current folder
structure on the shared epidemiology (`E:`) and data (`G:`) drives. The
`paths` list contains references to commonly used folders. These values
are environment-specific rather than portable defaults. If a drive is
remapped or a folder is renamed or moved, the corresponding `paths`
entry must be updated before dependent workflows will work correctly.

| Name | Folder | Data Types |
|:---------------|:----------------------------|:--------------------------|
| data | G:/Data/Communicable Disease/Current Year | Exported CSV files from EpiTrax and NBS |
| communicable | G:/Data/Communicable Disease | Parent folder for the `data` folder |
| scripts | G:/Data/Communicable Disease/R Scripts | R scripts used with exported data |
| cities | G:/Data/GIS/Shape Files/Tarrant/Cities | GIS polygon file for Tarrant Cities |
| spatial | G:/Data/GIS/Shape Files | General GIS files with several subfolders |
| population | G:/Data/Population | Census, projections, and estimates files |
| monthReportCode | E:/Monthly Report/Code | The folder for the monthly report R project |
| monthReport | E:/Monthly Report | Subfolders for each month's report |
| arbovirus | E:/Arboviral Diseases | Arboviral disease data files and guidance |
| immunization | E:/Immunization | Immunization data |
| flu | PublicHealth/Shared/EPI/Influenza | Influenza surveillance folder |

Here is an example of using `paths` to list files and folders in the
spatial-data folder:

```{r paths, eval=TRUE}
# The path to the data folder
paths$spatial

# List the file names in the spatial folder
files <- list.files(path = paths$spatial)  # Search for all files
head(files)  # View the first six in the folder
rm(files)
```

## Miscellaneous helpers

Several utility functions are included. These retrieve MMWR weeks and
dates, read and write tabular clipboard data, manipulate character
strings, and detect and convert yes/no vectors.

### MMWR weeks

A set of functions retrieves MMWR weeks (also called epidemiological
weeks) and their dates. The basic `mmwrWeek(date)` function returns the
numeric MMWR week for a date vector. Conversely,
`mmwrWeekBegin(weekNum, yr)` returns the first date of an MMWR week,
`mmwrWeekEnd(weekNum, yr)` returns its last date, and
`mmwrWeekMonth(weekNum, yr, label, abbr)` returns its month.

You may use either a `Date` or a `POSIXct` datetime to retrieve the MMWR
week number. This is useful if you want to calculate the MMWR week for
dates like onset, birth, report, diagnosis, etc. Here's an example of
using onset and getting the week number.

```{r mmwr_week, eval=TRUE}

# sample data
data("synthetic_outbreak", package = "tarr")
foo_flu <- synthetic_outbreak |> 
           dplyr::filter(condition == "fooflu") 

# add the MMWR Week number when onset occurred
foo_flu$mmwr_week <- mmwrWeek(foo_flu$onset)

# look at onset dates and epi week for the first few records
foo_flu[, c("onset", "mmwr_week")] |> head()

# table by week
janitor::tabyl(foo_flu$mmwr_week)
```

You may wish to convert an MMWR week to an actual date in a given year.
Two functions can do that: `mmwrWeekBegin()` returns the Sunday of the
week, and `mmwrWeekEnd()` returns the Saturday. To obtain the midweek
date, add three days to the value returned by `mmwrWeekBegin()`.

In addition to the week dates, `mmwrWeekMonth()` can be used
to get the month number, full name, or abbreviation. By default it
returns the month number for a given `weekNum` and `yr`. When the `label`
argument is set to TRUE, the function returns the full month name. When
the `label` and `abbr` arguments are both set to `TRUE`, the month
abbreviation is returned.

Using the date instead of the week number can help in interpreting
charts as a person does not have to mentally calculate the month when an
event occurs. The ability to convert a week number to a date in a
specific year also allows you to show data from different years using the
same date scale.

```{r mmwr_begin_end, eval=TRUE}
# show dates and month information from mmwrWeek
foo_flu <- foo_flu |> 
  head() |> 
  dplyr::mutate(
    mmwr_year = lubridate::epiyear(onset),
    begin_date = mmwrWeekBegin(weekNum = mmwr_week, yr = mmwr_year),
    end_date = mmwrWeekEnd(weekNum = mmwr_week, yr = mmwr_year),
    month_num = mmwrWeekMonth(mmwr_week, mmwr_year),
    month_name = mmwrWeekMonth(mmwr_week, mmwr_year, label = TRUE)
  )

foo_flu[, c("mmwr_week", "begin_date", "end_date", "month_num", "month_name")] |>
  head()
```

### Read and write to the clipboard

Two functions are provided to interact with the clipboard; `read.clip()`
and `write.clip(x)`. Both act on tabular data like a data frame and are
simple wrappers around `clipr::read_clip_tbl()` and
`clipr::write_clip()`. Passing a data frame to `write.clip()` places a
copy on the clipboard, which can then be pasted into another program,
such as Excel or Word. To move tabular data from Excel into R, highlight
the data including its headers, press **Ctrl+C**, and then assign the
clipboard data with `df <- read.clip()`.

### Special string functions

Several simple character string functions are provided to extract
classes of characters. They are collectively named "mash" or "only",
indicating the functions return only certain character types and can
result in the character string being "mashed." They are useful when
you want to have only alphanumeric characters in a string to improve
string matching.

The functions `only_digits(string)` and `mash_phone(string)` are
synonymous, returning only digit characters.

```{r mash_phone, eval=TRUE}

# phone numbers formatted differently
phone_numbers <- c("817/321-4700", "817-321-2366", "(214)972-9840")

# show only the digits
mash_phone(phone_numbers)

# or using the synonymous function
only_digits(phone_numbers) 
rm(phone_numbers)
```

The `only_alpha()` and `mash_name()` functions are also synonymous,
returning only alphabetic characters.

```{r only_alpha, eval=TRUE}
addr_roads <- c(
  "5001 Riverside",
  "1101 South Main Street",
  "Interstate 35",
  "Highway 287",
  "Hwy 10",
  "200 Weatherford Street"
)
  
# only alphabetic characters
mash_name(addr_roads)
only_alpha(addr_roads)

rm(addr_roads)
```

### Character to logical vectors

In many data sets, character strings containing yes/no values or
integers containing `0`/`1` are used in place of `FALSE`/`TRUE`.
Converting these vectors to R logical vectors can simplify analysis.
The following functions support this conversion:

- `is_yn(vec)` returns `TRUE` when every observed, nonmissing value is a
  recognized yes, no, or unknown code and the vector contains at least
  one yes or no value. Matching is case-insensitive. Numeric `1` means
  true and `0` means false. A `TRUE` result has a `values` attribute: a
  named list containing the observed codes normalized to lowercase.

- `id_yn_cols(df)` identifies the yes/no columns in a data frame and
  returns the names of columns that can be converted to logical vectors.

- `yn_2_logical(vec)` converts a recognized yes/no vector to a logical
  vector. The codes `u`, `unk`, `unknown`, `na`, blank strings, and
  missing values are converted to `NA`.

```{r yn_funcs, eval=TRUE}

set.seed(2026)
recovered <- sample(
  x = c("Yes", "No", "Unk"),
  size = 15,
  replace = TRUE,
  prob = c(0.50, 0.30, 0.20)
)
print(recovered)

is_yn(recovered)

# the original and converted vectors together in a data frame
rec_lgl <- yn_2_logical(recovered)
rec_df <- data.frame("recover_string" = recovered, "recover_logical" = rec_lgl)

# show the data frame 
rec_df

rm(recovered, rec_lgl, rec_df)

```

### General vector conversion function

It is common to have two vectors of different types when they need to
share the same type. R provides conversion functions such as
`as.numeric()`, `as.character()`, and `as.logical()`. The
`convert_type(vec1, vec2)` function generalizes this approach by using
`vec2` as a type template. It attempts to convert `vec1` to the type of
`vec2` when R supports that conversion.

```{r type_conversion, eval=TRUE}

# Dates represented as character string all in year-month-day format
char_dates <- c("2022-05-15", "2016/12/25", "1983-06-25")
date <- as.Date("1990-04-24")

converted <- convert_type(vec1 = char_dates, vec2 = date)
converted
class(converted)
```
